### 2025-10-05 thesis/design/indexing.tex
Better design section

In order to determine structural similarity of database schemas we propose using
graph representation of database schemas as a data definition language (DDL)
independent representation of the databases structure. Furthermore choosing a graph
representation allows us to use prevalent methods for measuring graph similarity
(eg, \textcolor{red}{Weisfeiler-Lehman graph kernels, cite}).

We define the structural similarity of two databases as the toplogical distance
of the respective \textsc{Natural}-graphs. The \textsc{Natural}-graph representation
only models tables, columns, relationships, datatypes and constraints, omitting 
table names, column names and domain terminology.

\begin{figure}[ht]
  \vspace{1em}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}] 
      \node[main] (1) {$t_1$}; 
      \node[main] (2) [below of=1] {$c_1$}; 
      \node[main] (3) [right of=2] {$t_2$}; 
      \node[main] (4) [below left of=3] {$c_3$}; 
      \node[main] (5) [below right of=3] {$c_4$};

      \draw (1) -- (2); 
      \draw (2) -- (4); 
      \draw (3) -- (4); 
      \draw (3) -- (5); 
      \draw (1) to [out=0, in=90, looseness=1] (3); 
    \end{tikzpicture} 
    \caption{Normalized graph repr.}
    \label{figure:normalized-graph-representation}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \begin{tikzpicture}[node distance={15mm}, main/.style = {draw, circle}] 
      \node[main] (1) {$t_1$}; 
      \node[main] (2) [below of=1] {$c_1$}; 
      \node[main] (3) [below right of=1] {$c_2$};

      \draw (1) -- (2); 
      \draw (1) -- (3); 
    \end{tikzpicture} 
    \caption{Denormalized graph repr.}
    \label{figure:denormalized-graph-representation}
  \end{minipage}
  \hfill
  \vspace{1em}
\end{figure}

Now measuring the distance between the graphs displayed in~\ref{figure:normalized-graph-representation} and~\ref{figure:denormalized-graph-representation}

% TODO: Finish this section on WWL graph kernels and distance calculation..

%is a set of id and distance tuples $\{(s_1, d_1), ..., (s_i,d_i)\}$
%where $s$ is the database schema name and $d$ is the distance to the
%\textit{current} schema. Thus a schema distance index has to be computed for each database schema
%\textsc{Natural} is running on.

%Thus.. $\mathcal{D}$

>>>>

To determine structural similarity of database schemas, we propose using
graph representation of database schemas as a data definition language (DDL)
independent representation of the database structure. Choosing a graph
representation allows us to leverage established methods for measuring graph similarity
such as Wasserstein-Weisfeiler-Lehman graph kernels \cite{WWL}. % TODO: Add proper WWL citation

Given a database schema $s$, the corresponding graph $G_s = (V, E, \ell, w)$ is defined as:

\begin{enumerate}
    \item $V = V_t \cup V_c$ where $V_t$ represents table nodes and $V_c$ represents column nodes
    \item $E = E_{tc} \cup E_{fk} \cup E_{ref}$ where:
        \begin{itemize}
            \item $E_{tc}$ are table-column edges
            \item $E_{fk}$ are foreign key relationship edges between tables
            \item $E_{ref}$ are reference edges between foreign key columns
        \end{itemize}
    \item Each node $v \in V$ has a semantic label $\ell(v) \in \{1, 2, \ldots, 9\}$ where:
        \begin{itemize}
            \item $\ell(v) = 1$ for table nodes
            \item $\ell(v) = 2$ for generic column nodes
            \item $\ell(v) = 3$ for primary key columns
            \item $\ell(v) = 4$ for foreign key columns
            \item $\ell(v) = 5$ for text columns
            \item $\ell(v) = 6$ for numeric columns
            \item $\ell(v) = 7$ for datetime columns
            \item $\ell(v) = 8$ for boolean columns
            \item $\ell(v) = 9$ for other/unknown data types
        \end{itemize}
    \item Each edge $e \in E$ has a weight $w(e) \in \{0.5, 0.9, 1.0\}$ reflecting structural importance:
        \begin{itemize}
            \item $w(e) = 1.0$ for foreign key relationships (highest importance)
            \item $w(e) = 0.9$ for column reference edges
            \item $w(e) = 0.5$ for table-column edges
        \end{itemize}
\end{enumerate}

We define the structural similarity of two databases as the topological distance
of the respective graphs. The graph representation captures essential schema
structure through semantic node labeling that prioritizes constraints over data
types: primary key columns receive label 3 regardless of their underlying data
type, foreign key columns receive label 4, and only then are remaining columns
categorized by data type (text=5, numeric=6, datetime=7, boolean=8, other=9).
This hierarchical labeling ensures that structural relationships take
precedence over semantic content, while omitting table names, column names and
domain terminology to achieve schema-agnostic comparison.

The graph representation of the database schemas described in~\ref{figure:schema-normalized}
and~\ref{figure:schema-denormalized} are therefore:

\begin{figure}[ht]
  \vspace{1em}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \begin{tikzpicture}[node distance={15mm}, table/.style = {draw,circle}, column/.style = {draw,dashed,circle}]

      \node[table] (users) at (0,0) {1}; 
      \node[table] (contacts) at (2,0) {1};

      \node[column] (users_id) [below left of=users] {3};

      \node[column] (contacts_user_id) [below left of=contacts] {4};
      \node[column] (contacts_name) [below right of=contacts] {5};

      \draw[dashed] (users) -- (users_id); 
      \draw[dashed] (contacts) -- (contacts_user_id); 
      \draw[dashed] (contacts) -- (contacts_name);

      \draw[thick] (users) to [out=10, in=170] (contacts);

      \draw (users_id) to [out=10, in=170] (contacts_user_id); 
    \end{tikzpicture} 
    \caption{Normalized graph repr.}
    \label{figure:normalized-graph-representation}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.45\linewidth}
    \centering
    \begin{tikzpicture}[node distance={15mm}, table/.style = {draw,circle}, column/.style = {draw,dashed,circle}]
      \node[table] (users) at (0,0) {1};

      \node[column] (users_id) [below left of=users] {3}; 
      \node[column] (users_contacts) [below right of=users] {9};

      \draw[dashed] (users) -- (users_id); 
      \draw[dashed] (users) -- (users_contacts); 
    \end{tikzpicture} 
    \caption{Denormalized graph repr.}
    \label{figure:denormalized-graph-representation}
  \end{minipage}
  \hfill
  \vspace{1em}

  \begin{minipage}{\linewidth}
    \centering
    \vspace{0.5em}
    \small
    \textbf{Legend:} Solid circle = table (1), dashed circle = column (3,4,5), 
    thick line = fk relationship ($w$=1.0), line = reference ($w$=0.9), 
    dashed line = table-column ($w$=0.5)
  \end{minipage}
\end{figure}

The $\delta$ of the order of the graphs~\ref{figure:normalized-graph-representation}
and~\ref{figure:denormalized-graph-representation}
further highlight the structural difference between the two schemas.

\subsubsection{Distance Measurement}

To measure the distance between the graphs displayed in~\ref{figure:normalized-graph-representation}
and~\ref{figure:denormalized-graph-representation}, we employ the
Wasserstein-Weisfeiler-Lehman (WWL) graph kernel method \cite{WWL}. % TODO: Fix WWL citation

The WWL kernel computes structural similarity by combining discrete Weisfeiler-Lehman
graph features with continuous optimal transport theory. The method first extracts
WL features from the labeled graph structure, then computes the Wasserstein distance
between the resulting feature distributions. For two \textsc{Natural}-graphs $G_1$ and $G_2$
with semantic node labels, the distance is computed as:

\begin{equation}
% TODO: Add missing WWL distance formula 
%d_{WWL}(G_1, G_2) = W_2(\mu_{G_1}, \mu_{G_2})
\end{equation}

where $W_2$ denotes the 2-Wasserstein distance and $\mu_{G_i}$ represents the empirical
measure over WL features extracted from graph $G_i$. The semantic node labels
$\ell(v) \in \{1, \ldots, 9\}$ guide the feature extraction process, ensuring that
structurally and semantically similar schema elements contribute similarly to the
distance calculation.

For the example schemas above, the normalized schema (5 nodes, 7 edges) and
denormalized schema (3 nodes, 2 edges) exhibit significant topological differences.
The WWL distance captures both the reduction in graph complexity (fewer nodes and edges)
and the loss of relational structure (elimination of foreign key relationships),
resulting in a substantial distance value reflecting their structural dissimilarity
despite representing equivalent logical relationships.

% TODO: Complete distance calculation example with actual values
% TODO: Add discussion of how semantic labeling affects distance calculation  
% TODO: Clarify relationship between edge weights and WWL computation
% TODO: Add example showing how label priority affects similarity scores

The schema distance index $\mathcal{D}$ maintains precomputed distances between
all observed database schemas, enabling efficient retrieval of structurally similar
databases during sample selection. This index is constructed as a set of
schema-distance tuples $\{(s_1, d_1), \ldots, (s_i, d_i)\}$ where $s_i$ represents
a database schema and $d_i$ denotes its topological distance to the current target
schema. The computational complexity of WL kernel calculation is $O(h \cdot |E|)$
where $h$ is the number of iterations and $|E|$ is the number of edges, making
it practical for real-time schema comparison in moderately sized databases.

% TODO: Fix inconsistent notation: use consistent variable names (G_s vs Natural-graphs)
% TODO: Improve transition between graph definition and distance measurement sections

### 2025-10-05 thesis/design/sampling.tex
Better design section

Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes 
cosine similarity in an embedding space $v$ to search for examples based on the
masked natural language query $q'$ and the respective masked SQL query $\omega'$.
This allows $\sigma$ to use cosine distance to search through historically observed
question answer pairs; eg, from previous user interactions or prevalent training
datasets like \textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.

@@

The natural language masking process replaces all words which are not deemed
\textit{structurally relevant} with the \texttt{<mask>} token. All words $w$
in a predefined whitelist $\mathcal{W}$ are determined structually relevant.
% TODO: Fix reference
Choosing a sound set $\mathcal{W}$ is described in section~\ref{implementation:whitelisting}.

@@

To construct an embedding space $\mathcal{V}_{\mathcal{S}}$ over a set of samples $\mathcal{S} = \{(q_1, s_1), ..., (q_i, s_i)\}$
where $q$ is the natural language question and $s$ is the corresponding SQL query, we need to compute
the embeddings of $mask(q)$ and $mask(s)$ respectively through an embedding function $\iota$.

@@

>>>>

Building upon DAIL-SQL's masked example selection, \textsc{Natural} utilizes 
cosine similarity in an embedding space $\mathcal{V}$ to search for examples based on the % TODO: Fix notation consistency with definitions
masked natural language query $q'$ and the respective masked SQL query $\omega'$.
This allows $\sigma$ to use cosine distance to search through historically observed
question answer pairs; eg, from previous user interactions or prevalent training
datasets like \textsc{Spider}, \textsc{Bird} or \textsc{SynSQL-2.5m}.

@@

The natural language masking process replaces all words which are not deemed
\textit{structurally relevant} with the \texttt{<mask>} token. All words $w$
in a predefined whitelist $\mathcal{W}$ are determined structurally relevant. % TODO: Fix typo
% TODO: Fix reference to implementation section
Choosing a sound set $\mathcal{W}$ is described in section~\ref{implementation:whitelisting}.

@@

To construct an embedding space $\mathcal{V}_{\mathcal{S}}$ over a set of samples $\mathcal{S} = \{(q_1, s_1), ..., (q_i, s_i)\}$
where $q$ is the natural language question and $s$ is the corresponding SQL query, we need to compute
the embeddings of $mask(q)$ and $mask(s)$ respectively through an embedding function $\iota$. % TODO: Specify embedding model

@@

% TODO: Explain how this relates to the selection function σ from definitions
% TODO: Discuss potential limitations of masked embedding approach

% ALIGNMENT IMPROVEMENTS NEEDED:
% TODO: Add formal mathematical definition using enumerate style like indexing section
% TODO: Use consistent tuple notation (q,s,ω,d) matching definitions section  
% TODO: Add cross-references to Section~\ref{design:indexing} for schema distances
% TODO: Include similarity scoring methodology with weights w_q, w_s, w_sql
% TODO: Add subsection on "Distance Calculation" parallel to indexing section
% TODO: Use consistent variable naming (mathcal{V} vs V vs v)
% TODO: Add example showing multi-dimensional scoring like indexing graphs
% TODO: Include legend/explanation of masking tokens like indexing legend
